### 2023-10-31 journal/correctness-GC.tex
editing Sec 6

If line \ref{appendLeafGC} of an \op{Append} operation modifies $\var{v}.\fld{blocks}$,
then \var{v} is a leaf node, and no other process ever modifies $\var{v}.\fld{blocks}$.
$T'$ was obtained from $T$ by calling \op{AddBlock}($T,B$).
$B$ was created either by the \op{Enqueue} or \op{Dequeue} that called \op{Append}.
Either way, $B.\fld{index} = \max(I)+1$.
The \op{AddBlock} that creates $T'$ may optionally \op{Split} the RBT and then add $B$ to it.
So the claim is satisfied.

@@

Since RBTs are always non-empty, calls to \op{MaxBlock} have well-defined answers.  
Throughout the proof, we use $\var{v}.\blocks[b]$ to refer to the block with \fld{index} $b$ that appeared
in \var{v}'s tree at some time during the execution.
It follows from Lemma \ref{RBTupdates} and Corollary \ref{nonEmptyIncreasing} that 
each time a new block appears in \var{v}'s RBT, its \fld{index} is greater than any block
that has appeared in \var{v}'s RBT earlier.
Thus, $\var{v}.\blocks[b]$ is unique, if it exists.
We also use this notation in the code to indicate that a search of the RBT should be performed for the block
with \fld{index} $b$.

@@

\begin{customlemma}{\ref{lem::headProgress}$'$}\label{lem::headProgressGC}
 If $h>0$ and a block with index $h$ has been inserted into $\var{v}.\fld{blocks}$ then 
 $\var{v}.\fld{blocks}[h-1].\fld{end\sub{left}} \leq \var{v}.\fld{blocks}[h].\fld{end\sub{left}}$ and 
 $\var{v}.\fld{blocks}[h-1].\fld{end\sub{right}} \leq \var{v}.\fld{blocks}[h].\fld{end\sub{right}}$.
\end{customlemma}
\begin{proof}
The block $B$ with index $h$ was installed into \var{v}'s RBT by the \op{CAS} at line \ref{casGC}.
Suppose that \op{CAS} changed the tree from $T$ to $T'$.
Before this \op{CAS}, line \ref{refreshReadTGC} read the tree $T$ from $\var{v}.blocks$,
line \ref{refreshReadMax} found a block $B'$  with \fld{index} $h-1$ in $T$,
and then line \ref{invokeCreateBlockGC} created the block $B$ with $\fld{index}=h$.
Since $B'$ was already in $T$ before $B$ was created, the
\op{CreateBlock}$(\var{v},b-1)$ that created $B'$ terminated before the 
\op{CreateBlock}$(\var{v},b)$ that created $B$ started.
By Corollary \ref{nonEmptyIncreasing}, the value that line \ref{createEndLeftGC} of \op{CreateBlock}$(\var{v},b-1)$ 
stores in $B'.\eleft$ is less than or equal to the value that line \ref{createEndLeftGC} 
of \op{CreateBlock}$(\var{v},b)$ stores in $B.\eleft$.  
Similarly, the values stored in $B'.\eright$ and $B.\eright$ at line \ref{createEndRightGC} satisfy the claim.
\end{proof}

Lemma \ref{lem::headProgressGC} implies that 
the nodes of an in-order traversal of any RBT have non-decreasing values of
$\eleft$ (and of $\eright$).
Thus, the searches for a block based on \eleft\ or \eright\ values 
at lines \ref{searchsuper3}, \ref{searchsuper1} and \ref{searchsuper2},
which are used to look for the superblock of a node or its predecessor, can be done in logarithmic time.

Lemma \ref{lem::subblocksDistinct}, Corollary \ref{cor::noDuplicates}, Invariant \ref{inv::sum} and Corollary \ref{blockNotEmpty} all 
hold for the modified algorithm.  Their proofs are identical to those given in 
Section \ref{sec::basicProperties} since they  depend only on Lemma \ref{lem::headProgress} (which can be replaced by Lemma \ref{lem::headProgressGC})
and the definition of subblocks given in (\ref{defsubblock}).
In particular, Invariant \ref{inv::sum} says that nodes in an in-order traversal of a RBT have non-decreasing
values of \fld{sum\sub{enq}}  so the searches of a RBT based on
\fld{sum\sub{enq}} values in lines
\ref{FRsearchGC} and \ref{getChildGC} can be done in logarithmic time.

@@

Next, we prove an analogue of Lemma \ref{successfulRefresh}.
We say a node \var{v} \emph{contains} an operation
if some block containing the operation has previously appeared in \var{v}'s RBT (even if 
the block has been removed from the RBT by a subsequent \op{Split}).

@@

Suppose $R$'s call to \op{CreateBlock} returns a new block $B$
that is installed in $\var{v}.\fld{blocks}$
by $R$'s \op{CAS} at line \ref{casGC}.
The \op{CreateBlock} set $B.\eleft$ to the maximum \fld{index} in $\var{v}.\fld{left}$'s RBT at line \ref{createEndLeftGC}.
By Corollary \ref{nonEmptyIncreasing}, this maximum \fld{index} is bigger than $i$.
By the definition of subblocks, some block in \var{v} contains $B_\ell$ as a subblock
and therefore \var{v} contains $op$.

@@

\begin{customlemma}{\ref{lem::doubleRefresh}$'$}\label{lem::doubleRefreshGC}
Consider two consecutive terminating calls $R_1$, $R_2$ to \op{Refresh}(\var{v}) by the same process.
All operations contained \var{v}'s children when $R_1$ begins
are contained in \var{v} when $R_2$ terminates.
\end{customlemma}
\begin{proof}
If either $R_1$ or $R_2$ performs a successful \op{CAS} at line \ref{casGC} or terminates at line \ref{addOPGC}, the claim follows
from Lemma \ref{successfulRefreshGC}.
So suppose both $R_1$ and $R_2$ perform a failed \op{CAS} at line \ref{casGC}.
Then some other \op{CAS} on $\var{v}.\fld{blocks}$ succeeds between the time each \op{Refresh}
reads $\var{v}.\fld{blocks}$ at line \ref{refreshReadTGC} and performs its \op{CAS} at line \ref{casGC}.
Consider the \op{Refresh} $R_3$ that does this successful \op{CAS} during $R_2$.
$R_3$ must have read $\var{v}.\fld{blocks}$ after the successful \op{CAS} during $R_1$.
The claim follows from Lemma \ref{successfulRefreshGC} applied to $R_3$.
\end{proof}

@@

\here{Technically, following proof might have to be wrapped up in a big induction with the argument
that outputs are linearized correctly, to argue that the matchup between enqueues and dequeues
is the same as it would be in the linearization ordering.  (Check whether this is really needed.)
Or alternatively, we could add an assumption that if every operation agrees with the linearization
so far, then the invariant is preserved by each step.
Then, when proving the linearization is correct, everything will work out.
But just leave it as is for the submission, since this is a subtle point}

@@

We show that every step preserves the invariant.  
We need only consider a step $st$ that modifies a node's RBT.
The minimum \fld{index} of \var{v}'s RBT can only change when \var{v}'s RBT changes, either at line 
\ref{appendLeafGC} (if \var{v} is a leaf) or at line \ref{casGC} (if \var{v} is an internal node).
In either case, the step $st$ changes $\var{v}.\fld{blocks}$ 
from $T$ to $T'$, where $T'$ is obtained by  a call $A$ to \op{AddBlock}(\var{v}, $T$, $B$).
(In the case of a leaf \var{v}, this is true because only the process that owns the leaf ever
writes to $\var{v}.\fld{blocks}$.)
If $A$ does not do GC (lines \ref{GCstart}--\ref{GCend}), 
then $T'$ is obtained by adding a new block to $T$, so by Lemma \ref{RBTupdates} the minimum \fld{index}
is unchanged and the invariant is trivially preserved.
So consider the case where $A$ performs GC.
We must show that any block of \var{v} whose index is less than or equal to the
minimum \fld{index} in $T'$ is finished when $T'$ is installed in $\var{v}.\fld{blocks}$.
Since $T'$ is obtained by discarding all blocks with \fld{index} at most $s$ (and adding
a block with a larger index), it suffices to show that all blocks that were ever added
to \var{v}'s RBT with \fld{index} at most $s$ are finished.

We must examine how $A$'s call to the recursive algorithm \op{SplitBlock} computes the value of $s$.

\begin{claim}\label{splitfinished}
If one of the recursive calls to $\op{SplitBlock}(x)$ within $A$'s call to \op{SplitBlock},
then that block and all earlier blocks in $x$ are finished when $st$ occurs.
\end{claim}
\begin{proof}[Proof of Claim]
We prove this claim  by induction on the depth of $x$.

For the base case, suppose $x$ is the root.
\op{SplitBlock} finds the maximum value $m$ in \var{last}, which is the index of some block
that contains an operation that is either a null dequeue or an enqueue whose value is the response for a dequeue that has been propagated to the root (since these are the only ways that an entry of \var{last} can be set to $m$).
By the FIFO property of queues, the values enqueued by enqueues in $\var{root}.\fld{blocks}[1..m-1]$
are all dequeued by operations that have already been installed in $\var{root}.\fld{blocks}$ before
the end of the \op{SplitBlock}.
Between the termination of $A$'s call to \op{SplitBlock} at line \ref{callSplitBlock}
and the \op{CAS} step $st$ after $A$ terminates,  $A$ helps all pending dequeues at line \ref{invokeHelp}.
Thus, after this helping (and before step $st$),
$\var{root}.\fld{blocks}[1..m-1]$ are all finished blocks, so the claim is true.

>>>>

If line \ref{appendLeafGC} of an \op{Append} operation modifies $\var{v}.\fld{blocks}$,
then \var{v} is a leaf node, and no other process ever modifies $\var{v}.\fld{blocks}$.
$T'$ was obtained from $T$ by calling \op{AddBlock}($v,T,B$).
$B$ was created either by the \op{Enqueue} or \op{Dequeue} that called \op{Append}.
Either way, $B.\fld{index} = \max(I)+1$.
The \op{AddBlock} that creates $T'$ may optionally \op{Split} the RBT at line \ref{splitGC} and then add $B$ to it.
So the claim is satisfied.

@@

Since RBTs are always non-empty, calls to \op{MaxBlock} have well-defined answers.  
Throughout the proof, we use $\var{v}.\blocks[b]$ to refer to the block with \fld{index} $b$ that appeared
in \var{v}'s tree at some time during the execution.
It follows from Lemma \ref{RBTupdates} and Corollary \ref{nonEmptyIncreasing} that 
each time a new block appears in \var{v}'s RBT, its \fld{index} is greater than any block
that has appeared in \var{v}'s RBT earlier.
Thus, $\var{v}.\blocks[b]$ is unique, if it exists.
We also use this notation in the code to indicate that a search of the RBT $v.\fld{blocks}$ should be performed for the block
with \fld{index} $b$.

@@

\begin{customlemma}{\ref{lem::headProgress}$'$}\label{lem::headProgressGC}
 If $v$ is an internal node and a block with index $h>0$ has been inserted into $\var{v}.\fld{blocks}$ then 
 $\var{v}.\fld{blocks}[h-1].\fld{end\sub{left}} \leq \var{v}.\fld{blocks}[h].\fld{end\sub{left}}$ and 
 $\var{v}.\fld{blocks}[h-1].\fld{end\sub{right}} \leq \var{v}.\fld{blocks}[h].\fld{end\sub{right}}$.
\end{customlemma}
\begin{proof}
The block $B$ with index $h$ was installed into \var{v}'s RBT by the \op{CAS} at line \ref{casGC}.
Suppose that \op{CAS} changed the tree from $T$ to $T'$.
Before this \op{CAS}, line \ref{refreshReadTGC} read the tree $T$ from $\var{v}.blocks$,
line \ref{refreshReadMax} found a block $B'$  with \fld{index} $h-1$ in $T$,
and then line \ref{invokeCreateBlockGC} created the block $B$ with $\fld{index}=h$.
Since $B'$ was already in $T$ before $B$ was created, the
\op{CreateBlock}$(\var{v},b-1)$ that created $B'$ terminated before the 
\op{CreateBlock}$(\var{v},b)$ that created $B$ started.
By Corollary \ref{nonEmptyIncreasing}, the value that line \ref{createEndLeftGC} of \op{CreateBlock}$(\var{v},b-1)$ 
stores in $B'.\eleft$ is less than or equal to the value that line \ref{createEndLeftGC} 
of \op{CreateBlock}$(\var{v},b)$ stores in $B.\eleft$.  
Similarly, the values stored in $B'.\eright$ and $B.\eright$ at line \ref{createEndRightGC} satisfy the claim.
\end{proof}

Lemma \ref{lem::headProgressGC} implies that 
the nodes of an in-order traversal of any RBT have non-decreasing values of
$\eleft$ (and of $\eright$).
Thus, the searches for a block based on \eleft\ or \eright\ values 
at lines \ref{searchsuper3}, \ref{searchsuper1} and \ref{searchsuper2},
which are used to look for the superblock of a node or its predecessor, can be done 
using an ordinary BST search.

Lemma \ref{lem::subblocksDistinct}, Corollary \ref{cor::noDuplicates}, Invariant \ref{inv::sum} and Corollary \ref{blockNotEmpty} all 
hold for the modified algorithm.  Their proofs are identical to those given in 
Section \ref{sec::basicProperties} since they  depend only on Lemma \ref{lem::headProgress} (which can be replaced by Lemma \ref{lem::headProgressGC})
and the definition of subblocks given in (\ref{defsubblock}).
In particular, Invariant~\ref{inv::sum} says that nodes in an in-order traversal of a RBT have non-decreasing
values of \fld{sum\sub{enq}}  so the searches for a block based on
\fld{sum\sub{enq}} values in lines
\ref{FRsearchGC} and \ref{getChildGC} can be done using an ordinary BST search.

@@

Next, we prove an analogue of Lemma \ref{successfulRefresh}.
We say a node \var{v} \emph{contains} an operation
if some block containing the operation has previously appeared in the RBT \var{v}.\fld{block}
(even if the block has been removed from the RBT by a subsequent \op{Split} during garbage collection).

@@

Suppose $R$'s call to \op{CreateBlock} returns a new block $B$
that is installed in $\var{v}.\fld{blocks}$
by $R$'s \op{CAS} at line \ref{casGC}.
The \op{CreateBlock} set $B.\eleft$ to the maximum \fld{index} in $\var{v}.\fld{left}$'s RBT at line \ref{createEndLeftGC}.
By Corollary \ref{nonEmptyIncreasing}, this maximum \fld{index} is bigger than $i$.
By the definition of subblocks, some block in \var{v} contains $B_\ell$ as a subblock
and therefore \var{v} contains~$op$.

@@

\begin{customlemma}{\ref{lem::doubleRefresh}$'$}\label{lem::doubleRefreshGC}
Consider two consecutive terminating calls $R_1$, $R_2$ to \op{Refresh}(\var{v}) by the same process.
All operations contained \var{v}'s children when $R_1$ begins
are contained in \var{v} when $R_2$ terminates.
\end{customlemma}
\begin{proof}
If either $R_1$ or $R_2$ performs a successful \op{CAS} at line \ref{casGC} or terminates at line \ref{addOPGC}, the claim follows
from Lemma \ref{successfulRefreshGC}.
So suppose both $R_1$ and $R_2$ perform a failed \op{CAS} at line \ref{casGC}.
Then some other \op{CAS} on $\var{v}.\fld{blocks}$ succeeds between the time each \op{Refresh}
reads $\var{v}.\fld{blocks}$ at line \ref{refreshReadTGC} and performs its \op{CAS} at line \ref{casGC}.
Consider the \op{Refresh} $R_3$ that does this successful \op{CAS} during $R_2$.
$R_3$ must have read $\var{v}.\fld{blocks}$ after the successful \op{CAS} during $R_1$.
The claim follows from Lemma \ref{successfulRefreshGC} applied to~$R_3$.
\end{proof}

@@

%\here{Technically, following proof might have to be wrapped up in a big induction with the argument
%that outputs are linearized correctly, to argue that the matchup between enqueues and dequeues
%is the same as it would be in the linearization ordering.  (Check whether this is really needed.)
%Or alternatively, we could add an assumption that if every operation agrees with the linearization
%so far, then the invariant is preserved by each step.
%Then, when proving the linearization is correct, everything will work out.
%But just leave it as is for the submission, since this is a subtle point}
% NO, I DON'T THINK THIS IS NECESSARY

@@

We show that every step preserves the invariant.  
We need only consider a step $st$ that modifies a node's RBT.
The minimum \fld{index} of \var{v}'s RBT can only change when \var{v}'s RBT changes, either at line 
\ref{appendLeafGC} (if \var{v} is a leaf) or at line \ref{casGC} (if \var{v} is an internal node).
In either case, the step $st$ changes $\var{v}.\fld{blocks}$ 
from $T$ to $T'$, where $T'$ is obtained by  a call $A$ to \op{AddBlock}(\var{v}, $T$, $B$).
(In the case of a leaf \var{v}, this is true because only the process that owns the leaf ever
writes to $\var{v}.\fld{blocks}$.)
If $A$ does not do GC (lines \ref{GCstart}--\ref{GCend}), 
then $T'$ is obtained by adding a new block to $T$, so by Lemma \ref{RBTupdates} the minimum \fld{index}
is unchanged and the invariant is trivially preserved.
So consider the case where $A$ performs GC.
We must show that any block of \var{v} whose index is less than or equal to the
minimum \fld{index} in $T'$ is finished when $T'$ is installed in $\var{v}.\fld{blocks}$.
Since $T'$ is obtained by discarding all blocks with \fld{index} values less than or equal to $s$ (and adding
a block with a larger index), it suffices to show that all blocks that were ever added
to \var{v}'s RBT with \fld{index} at most $s$ are finished.

We must examine how $A$'s call at line \ref{callSplitBlock} to the recursive algorithm \op{SplitBlock} computes the value of $s$.

\begin{claim}\label{splitfinished}
If one of the recursive calls to $\op{SplitBlock}(x)$ within $A$'s call to \op{SplitBlock} returns a block $B$,
then $B$ and all earlier blocks in $x$ are finished when $st$ occurs.
\end{claim}
\begin{proof}[Proof of Claim]
We prove this claim  by induction on the depth of $x$.

For the base case, suppose $x$ is the root.
\op{SplitBlock} finds the maximum value $m$ in \var{last}, which is the index of some block
that contains an operation that is either a null dequeue or an enqueue whose value is the response for a dequeue that has been propagated to the root (since these are the only ways that an entry of \var{last} can be set to $m$).
By the FIFO property of queues, the values enqueued by enqueues in $\var{root}.\fld{blocks}[1..m-1]$
are all dequeued by operations that have already been installed in $\var{root}.\fld{blocks}$ before
the end of the \op{SplitBlock}.
Between the termination of $A$'s call to \op{SplitBlock} at line \ref{callSplitBlock}
and the \op{CAS} step $st$ after $A$ terminates,  $A$ helps all pending dequeues at line \ref{invokeHelp}.
Thus, after this helping (and before step $st$),
$\var{root}.\fld{blocks}[1..m-1]$ are all finished blocks.
Since \op{SplitBlock} returns \var{root}.\fld{blocks}[$m-1$], the claim is true.
